---
output: rmarkdown::html_vignette
author: Fabian Mueller
date: "`r Sys.Date()`"
title: ChrAccR
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`ChrAccR` is an R package for the analysis of chromatin accessibility data.
It supports popular protocols like the Assay for Transposase-Accessible Chromatin
(ATAC-seq) and Nucleosome Occupancy and Methylome Sequencing (NOMe-seq; beta-stage)
in bulk and single-cell scenarios.

## Installation

Use library or require to load the package and useful additional packages.

```{r, message = FALSE, warning = FALSE, eval = FALSE}
# install devtools if not previously installed
if (!is.element('devtools', installed.packages()[,"Package"])) install.packages('devtools')

# install dependencies
devtools::install_github("demuellae/muLogR")
devtools::install_github("demuellae/muRtools")

# install ChrAccR
devtools::install_github("demuellae/ChrAccR")
```

## Loading
```{r, message = FALSE, warning = FALSE}
library(ChrAccR)
```
